Supplementary Information

### 2021-07-30 doc/manuscript/ForestGEO_treerings.Rmd
work on responses to reviewers
- DBH measurements
- detrending methods
add first draft of summary figure
#136

#Figures
schematic=1
SCBI_example=2
clim_sens=3
interactions_fig = 4 #DBH/climate interactions
size=5
year_fig = 6

@@

We used the *climwin* package in R [@vandepol_identifying_2016] to identify the most important climate variable and the seasonal window over which its effect was strongest for each of two categories of variables: a temperature group ($T_{min}$, $T_{max}$, $T_{mean}$, and PET) and a precipitation group (PPT, PDF).
To remove low-frequency variation that most likely represents responses to non-climatic drivers (*e.g.*, growth and aging of the tree, change in competitive dynamics, atmospheric pollution), we detrended the response variables by fitting penalized thin-plate regression splines in generalized additive models [GAM, functions *gam* and *s* in the R package *mgcv*, @wood_fast_2011] to individual growth records (RW, BAI, or $\Delta$AGB) from each core, and extracted the residual variation for each observation. 
The smoothing parameters were automatically selected by the *gam* function with generalized cross-validation (GCV).
We then used *climwin* to identify the climate drivers that most strongly correlated with the individual tree-level residuals of the growth variables, RW, BAI, or $\Delta$AGB, specifying linear and quadratic terms to allow for potential nonlinearities in the climate response.
Within *climwin*, we specified a mixed-effects model in which the fixed effects were the climate variables and the random intercepts were species (when n $\ge$ 3) and core identity (noting that these effects should be minimal given that residuals are centered around zero).
For each climate variable, we ran permutations for all possible combinations of consecutive months within a 15-month period ending near the time of cessation of formation of each annual ring (Table `r site_details`). 
*Climwin* runs all potential models to select the best fit (lowest AIC), and does k-fold cross-validation in its computation of AIC to guard against over-fitting [@vandepol_identifying_2016]. 
For each group of candidate climate variables (water and temperature; Fig. `r schematic`), we selected the variable - seasonal window combination with the lowest AIC as a candidate climate variable for the multivariate models.

@@

@@

![**Figure 3 | Species-level responses of RW to climwin-selected variables in precipitation and temperature variable groups.** Primary climate drivers are coded on the x-axes as the climate variable abbreviation followed by the range of months (p=previous year, c=current year) over which it is most influential. For each species (color-coded as in Fig. 5), relationships are plotted if included in the top model. For each relationship shown, other terms in the model are held constant at their medians. Best-fit polynomials are plotted with solid lines when both first- and second-order terms are significant (t-test’s p-value <0.05), dash-dotted lines when only one term is significant, and dotted lines when neither is significant. Transparent ribbons indicate 95% confidence intervals. Vertical grey lines indicate the long-term mean for the climate driver over the analysis period; shading indicates 1 SD.](tables_figures/pre_temp_groups.png)

@@

![**Figure 4 | Examples of climate - DBH interactions for three species at three sites.** Shown are modeled response functions at the minimum and maximum and maximum tails of the DBH distribution. Other terms in the model are held constant at their medians. Transparent ribbons indicate 95% confidence intervals. Vertical grey lines indicate the long-term mean for the climate driver over the analysis period; shading indicates 1 SD.](tables_figures/pre_temp_groups_dbh_interactions.png)

@@

![**Figure 5 | Growth sensitivity to DBH: (a) RW, (b) BAI, (c) $\Delta$AGB.** Relationships for species are plotted when included in the top model. Other terms in the model are held constant at their medians. Best-fit polynomials are plotted with solid lines when both first- and second-order terms are significant (t-test’s p-value <0.05), dash-dotted lines when only one term is significant, and dotted lines when neither is significant.  Transparent ribbons indicate 95% confidence intervals.](tables_figures/DBH_responses.png)

@@

![**Figure 6 | Effect of year, when included in the best model, on BAI.** For each species (all listed), relationships are plotted if the year effect could be analyzed and was included in the top model. Other terms in the model are held constant at their medians. Best-fit polynomials are plotted with solid lines when both first- and second-order terms are significant (t-test’s p-value <0.05), dash-dotted lines when only one term is significant, and dotted lines when neither is significant.  Transparent ribbons indicate 95% confidence intervals. ](tables_figures/Year_responses_BAI_only.png)

>>>>

#Figures
schematic=1
SCBI_example=2
SummaryFigure=3
clim_sens=4
interactions_fig = 5 
size=6
year_fig = 7

@@

We used the *climwin* package in R [@vandepol_identifying_2016] to identify the most important climate variable and the seasonal window over which its effect was strongest for each of two categories of variables: a temperature group ($T_{min}$, $T_{max}$, $T_{mean}$, and PET) and a precipitation group (PPT, PDF).
To remove low-frequency variation that most likely represents responses to non-climatic drivers (*e.g.*, growth and aging of the tree, change in competitive dynamics, atmospheric pollution), we detrended the response variables by fitting penalized thin-plate regression splines in generalized additive models [GAM, functions *gam* and *s* in the R package *mgcv*, @wood_fast_2011] to individual growth records (RW, BAI, or $\Delta$AGB) from each core, and extracted the residual variation for each observation. 
The smoothing parameters were automatically selected by the *gam* function with generalized cross-validation (GCV).
These types of detrending methods are used to maximize the climate signal in tree-ring paleoclimate reconstructions and thus have been shown to retain the climate signal while removing the influence of other non-climatic drivers [@fritts_tree_1976; @cook_time_1985].  
We then used *climwin* to identify the climate drivers that most strongly correlated with the individual tree-level residuals of the growth variables, RW, BAI, or $\Delta$AGB, specifying linear and quadratic terms to allow for potential nonlinearities in the climate response.
Within *climwin*, we specified a mixed-effects model in which the fixed effects were the climate variables and the random intercepts were species (when n $\ge$ 3) and core identity (noting that these effects should be minimal given that residuals are centered around zero).
For each climate variable, we ran permutations for all possible combinations of consecutive months within a 15-month period ending near the time of cessation of formation of each annual ring (Table `r site_details`). 
*Climwin* runs all potential models to select the best fit (lowest AIC), and does k-fold cross-validation in its computation of AIC to guard against over-fitting [@vandepol_identifying_2016]. 
For each group of candidate climate variables (water and temperature; Fig. `r schematic`), we selected the variable - seasonal window combination with the lowest AIC as a candidate climate variable for the multivariate models.

@@

## Full model results overview {-#Full}

Fig. `r SummaryFigure`

```{r , eval=FALSE, echo=FALSE}
# This draw.io diagram that can be edited online here: https://app.diagrams.net/#HEcoClimLab%2FForestGEO-tree-rings%2Fmaster%2Fdoc%2Fmanuscript%2Ftables_figures%2FSummaryFigure.drawio
```
![**Figure 3 | **(SUMMARY FIGURE)**.](tables_figures/SummaryFigure.png)

@@

![**Figure 4 | Species-level responses of RW to climwin-selected variables in precipitation and temperature variable groups.** Primary climate drivers are coded on the x-axes as the climate variable abbreviation followed by the range of months (p=previous year, c=current year) over which it is most influential. For each species (color-coded as in Fig. 5), relationships are plotted if included in the top model. For each relationship shown, other terms in the model are held constant at their medians. Best-fit polynomials are plotted with solid lines when both first- and second-order terms are significant (t-test’s p-value <0.05), dash-dotted lines when only one term is significant, and dotted lines when neither is significant. Transparent ribbons indicate 95% confidence intervals. Vertical grey lines indicate the long-term mean for the climate driver over the analysis period; shading indicates 1 SD.](tables_figures/pre_temp_groups.png)

@@

![**Figure 5 | Examples of climate - DBH interactions for three species at three sites.** Shown are modeled response functions at the minimum and maximum and maximum tails of the DBH distribution. Other terms in the model are held constant at their medians. Transparent ribbons indicate 95% confidence intervals. Vertical grey lines indicate the long-term mean for the climate driver over the analysis period; shading indicates 1 SD.](tables_figures/pre_temp_groups_dbh_interactions.png)

@@

![**Figure 6 | Growth sensitivity to DBH: (a) RW, (b) BAI, (c) $\Delta$AGB.** Relationships for species are plotted when included in the top model. Other terms in the model are held constant at their medians. Best-fit polynomials are plotted with solid lines when both first- and second-order terms are significant (t-test’s p-value <0.05), dash-dotted lines when only one term is significant, and dotted lines when neither is significant.  Transparent ribbons indicate 95% confidence intervals.](tables_figures/DBH_responses.png)

@@

![**Figure 7 | Effect of year, when included in the best model, on BAI.** For each species (all listed), relationships are plotted if the year effect could be analyzed and was included in the top model. Other terms in the model are held constant at their medians. Best-fit polynomials are plotted with solid lines when both first- and second-order terms are significant (t-test’s p-value <0.05), dash-dotted lines when only one term is significant, and dotted lines when neither is significant.  Transparent ribbons indicate 95% confidence intervals. ](tables_figures/Year_responses_BAI_only.png)
